Using Planning to Generate Natural-Language Sentences:\\ Experiences and Challenges

### 2009-12-16 trunk/Papers/crisp-ff-10/crisp-ff.tex
ff sec revision, teil 2

First, while pre-processing the ADL formulas, FF has a sub-routine
that checks the instantiated formulas -- quantifiers already compiled
into grounded conjunctions and disjunctions -- for tautologies. This
involves a loop checking all pairs of sub-formulas within every
conjunction and disjunction, testing whether they are
identical/identical modulo their sign. For example, if we find $\phi$
and $\neg \phi$ within a disjunction, then the disjunction is replaced
by TRUE. Now, we haven't tested thoroughly whether this sub-routine
has any notable effect on the formulas in other benchmarks. It
certainly has no such effect in our domain. The effect it {\em does}
have is take a huge amount of runtime on the long goal conjunctions
stating that we do not wish to have any open substitution nodes nor
any distractors. The simple fix is hence to switch this sub-routine
off. In smaller examples, this gives advantages like $0.29s$
vs.\ $11s$ pre-processing time; in a larger example we ran, original
FF's pre-process doesn't finish within 15 minutes whereas our fix
yields a pre-processing time of about $3s$.

Second, FF's pre-process contains a very small imprudence that
apparently didn't surface in other domains, but that does in ours. FF
distinguishes operators into ``easy'' (DNF precondition) and ``hard''
(non-DNF precondition). The easy ones are instantiated in a
spezialized very effectively implemented procedure, while the hard
ones go through a much less effectively implemented more general
procedure. In particular, that latter procedure uses a data structure
encompassing an integer for every possible instantiation of all
predicates. Since this does not take into account reachability, it is
prohibitively huge in our domain. Now, there actually are no hard
operators in our domain. But FF builds the data structure anyway, even
though it is not needed. In small examples, this results in a couple
of seconds runtime overhead. In larger examples, it exhausts memory
and causes FF to terminate with a segmentation fault.

@@

While the goal agenda and related techniques have proved quite useful
in many domains, in the sentence generation domain they are completely
off-target. Reasonable orders exist {\bf \dots explain using the
  example dedicated to this point how we get to have only the
  ``subst'' goal in the first entry \dots} When achieving this first
goal agenda entry, the planner decides to generate {\em any}
sentence. If it happens to be the wrong sentence, the planner is stuck
in a dead-end. While that dead end is recognized by the relaxed plan
heuristic -- i.e., it is easy to see that the actions needed for
achieving the goal cannot be applied anymore -- certainly this is an
entirely unintended (and a bit amusing) outcome of this
technique. Switching it off, we {\bf \dots example before/after.}

{\bf \dots pose as a challenge \dots if sensible based on example
  observations, speculate a bit how/whether goal ordering techniques
  could be modified to work}

@@

To give just one concrete example, say we are in the initial state,
and have not yet committed to a sentence. For certain kinds of
sentences ({\bf more concrete here???}), the relaxed plan has length
$1$ because the we only need to choose the desired meaning. However,
in the state $s$ resulting from this action, the mechanisms of the
planning task introduce all the sub-goals that are required in order
to pin-point that meaning (in particular we get a new goal for all the
distractors that must be removed). Hence the relaxed plan is much
longer than before -- arbitrarily longer, in general. For FF's
particular search algorithm, enforced hill-climbing (EHC), this means
that {\em FF performs a blind breadth-first search}: EHC starts at $h$
value $1$ and thereafter needs to solve the whole problem in order to
find a (goal) state with lower $h$ value.

@@

{\bf \dots pose as a challenge \dots if sensible based on example
  observations, speculate a bit how/whether heuristics could be
  modified to work \dots a question that we should ideally at least
  partially answer in here is whether other existing heuristics
  actually work better.. we could try at least a little bit with PDBs,
  merge\&shrink, etc}

The only FF technique that appears to work reasonably well is its
action pruning technique, the ``helpful actions''. While the relaxed
plan is often much too short, it appears to always contain at least
one of the actions that actually are useful, and it certainly prunes
the set of avaibale actions quite effectively. For example, in one of
the instances where EHC turns into a breadth-first search, $16875$
nodes are explored without helpful actions; with helpful actions, this
reduces to $3078$. For larger examples, the factor between the two
numbers tends to become larger. {\bf use here larger version of
  modifiers-10-10???}

>>>>

First, while pre-processing the ADL formulas, FF has a sub-routine
that checks the instantiated formulas -- quantifiers already compiled
into grounded conjunctions and disjunctions -- for tautologies. This
involves a loop checking all pairs of sub-formulas within every
conjunction and disjunction, testing whether they are
identical/identical modulo their sign. For example, if we find $\phi$
and $\neg \phi$ within a disjunction, then the disjunction is replaced
by TRUE. Now, we haven't tested thoroughly whether this sub-routine
has any notable effect on the formulas in other benchmarks. It
certainly has no such effect in our domain. The effect it {\em does}
have is take a huge amount of runtime on the long goal conjunctions
stating that we do not wish to have any open substitution nodes nor
any distractors. The simple fix is hence to switch this sub-routine
off. In small instances, this gives advantages like $0.29s$ vs.\ $11s$
pre-processing time; in a larger instance we ran, original FF's
pre-process doesn't finish within 15 minutes whereas our fix yields a
pre-processing time of about $3s$.

Second, FF's pre-process contains a very small imprudence that
apparently didn't surface in other domains, but that does in ours. FF
distinguishes operators into ``easy'' (DNF precondition) and ``hard''
(non-DNF precondition). The easy ones are instantiated in a
spezialized very effectively implemented procedure, while the hard
ones go through a much less effectively implemented more general
procedure. In particular, that latter procedure uses a data structure
encompassing an integer for every possible instantiation of all
predicates. Since this does not take into account reachability, it is
prohibitively huge in our domain. Now, there actually are no hard
operators in our domain. But FF builds the data structure anyway, even
though it is not needed. In small instances, this results in a couple
of seconds runtime overhead. In larger instances, it exhausts memory
and causes FF to terminate with a segmentation fault.

@@

While the goal agenda and related techniques have proved quite useful
in many domains, in the sentence generation domain they often are
completely off-target. The most striking point concerns the
interaction between the goal to create a sentence -- $\neg
\mathsf{subst}(S,\mathsf{root})$ in the rabbit example -- and the goal
to express a certain meaning -- $\mathsf{sleep}(e,r_1)$ in the
rabbit example. The goal agenda orders the former before the latter
because it detects that, once a meaning is expressed, the sentence
creation has started and it is not possible to start anew
anymore. Unless the inverse also holds,\footnote{The inverse holds if,
  in the given grammar, the only way to create a meaning is to create
  a full sentence -- i.e., there are no side sentences like ``the
  sleeping rabbit''.} $\mathsf{subst}(S,\mathsf{root})$ is alone in
the first goal agenda entry, i.e., there is no goal stating the actual
meaning the sentence is supposed to have. Consequently, the planner
decides to generate {\em any} sentence. If it happens to be the wrong
sentence, the planner is stuck in a dead-end. While that dead end is
recognized by the relaxed plan heuristic -- i.e., it is easy to see
that the actions needed for achieving the goal cannot be applied
anymore -- certainly this is an entirely unintended (and a bit
amusing) outcome of this technique. 
% Switching it off, we {\bf \dots
%example before/after.}

It is not clear to us, at this point, how goal ordering techniques
could be modified so that they return sensible results in the sentence
generation domain. The issue explained above regarding $\neg
\mathsf{subst}(S,\mathsf{root})$ and $\mathsf{sleep}(e,r_1)$ could be
addressed by recognizing that, once the latter is achieved, the former
is already true anyway. More generally however, the life-cycle of
substitution nodes and distractors is weird -- as we outline below, it
follows a false-true-false pattern -- and seems difficult to detect
based on domain analysis.

@@

Consider again the rabbit example. In the initial state, the relaxed
plan has length $1$ because we only need to achieve $\neg
\mathsf{subst}(S,\mathsf{root})$ and $\mathsf{sleep}(e,r_1)$, which is
done by the single action ``sleeps''. However, that action has the
effect ``forall (?y - individual) (when (not (= ?y ?xself))
(distractor ?subjnode ?y))'' which introduces new distractors. Those
must be removed, so they form ``new'' goals. In the given case, the
relaxed plan now contains two actions to achieve these goals. In
general (if there are $n$ rabbits, say), the relaxed plan may become
arbitrarily long.
%To give just one concrete example, say we are in the initial state,
%and have not yet committed to a sentence. For certain kinds of
%sentences ({\bf more concrete here???}), the relaxed plan has length
%$1$ because the we only need to choose the desired meaning. However,
%in the state $s$ resulting from this action, the mechanisms of the
%planning task introduce all the sub-goals that are required in order
%to pin-point that meaning (in particular we get a new goal for all the
%distractors that must be removed). Hence the relaxed plan is much
%longer than before -- arbitrarily longer, in general.  
For FF's particular search algorithm, enforced hill-climbing (EHC),
this means that {\em FF performs a blind breadth-first search}: EHC
starts at $h$ value $1$ and thereafter needs to solve the whole
problem in order to find a (goal) state with lower $h$ value.

@@

Can we design heuristics that are better at deadling with the
lifecycle of substitution nodes and distractors? The problem of not
noticing them in the initial state is present for all approaches that
are dominated by $h^+$,
e.g.\ \cite{karpas:domshlak-ijcai-09,helmert:domshlak:icaps-09}, and
is also present for the known inadmissible landmarks heuristics
\cite{richter:etal:aaai-08}. As far as known heuristics are concerned,
hope thus resides mainly in the $h^m$ family and in abstraction
heuristics. We ran a few tests with an example like the rabbit one,
except that there are 10 distractors. The Graphplan estimate for the
initial state is $4$, i.e., $h^2$ is better than $h^+$ here. Still,
$4$ is far away from the real goal distance $14$. With merge\&shrink
\cite{helmert:etal:icaps07}, {\bf \dots (malte agreed to run a test on
  modifiers-10-10-)}

The only FF technique that appears to work reasonably well is its
action pruning technique, the ``helpful actions''. While the relaxed
plan is often much too short, it appears to always contain at least
one of the actions that actually are useful, and it certainly prunes
the set of avaibale actions quite effectively. For example, in one of
the instances where EHC turns into a breadth-first search, $16875$
nodes are explored without helpful actions; with helpful actions, this
reduces to $3078$. For larger examples, the factor between the two
numbers tends to become larger. 
%{\bf use here larger version of
%  modifiers-10-10???}
